Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-04-14 doc/invt.tex
add is2.c

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry
that can provide insight into an array of phenomena not easily studied
with traditional experiments.
%
Given a system,
often the task is to compute
a distribution, $p^*(z)$,
along a collective variable $z$ by sampling via either Monte Carlo\cite{
  frenkel, newman, landau_binder} (MC)
or molecular dynamics\cite{frenkel, karplus2002} (MD) simulations.
%
The dimensionless free energy, or potential of mean force (PMF),
along the collective variable, $z$,
is $-\ln p^*(z)$.
%
However, accurately estimating the PMF is often difficult
when the collective variable is energetically restricted to local regions
of the complex, multi-modal free energy surface.
%
Thus,
to overcome these energetic barriers and
to capture the global shape of the free energy surface
an effective strategy is to introduce a bias potential that
cancels the PMF.
%
Ideally, then, sampling yields a flat distribution
along $z$\cite{mezei1987, berg1992, *lee1993,
wang2001, *wang2001pre,
huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}.
%
The flat-distribution sampling can be used to accelerate
sampling of slow transitions that otherwise would not readily
be observed with traditional simulations.
%
%A better solution is to artificially flatten
%the target distribution along $z$\cite{mezei1987, berg1992, *lee1993,
%wang2001, *wang2001pre,
%huber1994,
%*laio2002, *laio2008, *barducci2011, *sutto2012}.
%%
%This alleviates the above problem of local trapping
%by forcing the system to travel more
%deliberately along $z$,
%which often represents
%a direction of slow motion.
%%
%To achieve a flat distribution, however,
%the method must be able to construct a
%bias potential that cancels the PMF.

The Wang-Landau (WL) algorithm\cite{wang2001, *wang2001pre}
and metadynamics\cite{huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
flat-distribution-sampling (FDS) techniques
that build the bias potential adaptively or on the fly.
%
Both techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential.
%
The bias potential is associated with discrete values of $z$.
%
The difference lies
in the updating window function,
which takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one  has to gradually decrease
the updating magnitude of the bias.

@@

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule, $\alpha(t)$,
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014}.
%
An interesting explanation\cite{
  marsili2006, barducci2008}
comes from the interpretation of
the bias potential %under the WL updating scheme, Eq. \eqref{eq:wl_update},
as a runtime average from all data collected so far,
so that the addition of a new data point only
affects the average by a weight of $1/t$,
which is the inverse of the current sample size, $t$
(cf. Appendix \ref{sec:equilerr}).
%
Here we will re-derive this formula
and suggest a slight modification
whereby the origin of $t$ is shifted
according to the initial error.
%
Interestingly, we may reach the same result
from optimally planning the simulation lengths in the stages
and the reduction factor, as shown in Appendix \ref{sec:wltoinvt}.

@@

Note that the distribution, $\pmb \rho$,
introduced above to define the error
is independent of
the target distribution,
$\mathbf p = (p_1, \dots, p_n)$.
%
For example,
in well-tempered metadynamics,
$\pmb\rho$ is a flat distribution,
while $\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
barducci2008, dama2014}.
%
From Eqs. \eqref{eq:x_def} and \eqref{eq:vbar_def},
we have
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%

@@

%
For the random noise part,
the average is zero by definition,
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the autocorrelation functions
depend only on the time difference
%
\begin{equation}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  =
  \sigma_{ij}(t - t'),
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function
that vanishes at large $t$.
%
\note{More explicitly,
if the transition matrix of a period, $\tau$,
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \, \pi_j - \pi_i \, \pi_j.
$
}%
Below, we will assume the noise is white,
such that
$$
\sigma_{ij}(t) = \Gamma_0 \, \delta(t)
$$
for some $\Gamma_0$,
and $\delta(t)$ is the Dirac's delta function.
%
This approximation is valid if the simulation period, $T$,
is much longer than the autocorrelation time of $\sigma_{ij}(t)$,
%
and for a non-white noise,
the equivalent value of $\Gamma_0$ can be obtained
from the integral of the autocorrelation function,
$\int_{-\infty}^\infty \sigma_{ij}(t) \, dt$.

@@

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
where
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ j = 1 }^n
    \rho_j
    \frac{ \zeta_j(t) }
         {     p_j    }
  ,
\label{eq:xi_def}
\end{equation}
%
is also a white noise as $\zeta_i(t)$,
i.e. we may write
$$
  \langle \xi_i(t) \, \xi_i(t') \rangle = \Gamma_i \, \delta(t - t'),
$$
for some $\Gamma_i$.
%
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ h_j } { p_j }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} \; and \; \eqref{eq:xi_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ \langle h_j \rangle } { p_j }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_ave} \;and\; \eqref{eq:rhox_sum}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

@@

Consider the following stochastic differential equation
of a single variable:
%
\begin{equation}
\dot x(t) = -\alpha(t) \, \left[ x(t) - \xi(t) \right],
\label{eq:dxdt_alpha}
\end{equation}
%
where $\xi(t)$ is a white noise, such that
%
$
  \left\langle \xi(t) \, \xi(t') \right\rangle
  =
  \Gamma \, \delta(t - t')
  .
$
%
We wish to show that the $\alpha(t)$
of minimizing $\langle x^2(t) \rangle$ at long times
is given by
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 },
\label{eq:alpha_invt1}
\end{equation}
%
where $t_0$ is a constant.

@@

We can now simplify Eq. \eqref{eq:vt_diffeq_mbin}.
%
By using Eqs.
\eqref{eq:v_def}, \eqref{eq:vbar_def},
\eqref{eq:vt_diffeq_mbin}-\eqref{eq:w_balance}
we get
%
\begin{align*}
  \dot{ \bar v }(t)
  =
  \alpha(t)
  \sum_{r = 1}^n \rho_r
    \, \frac{ h_r(t) } { p_r }
  =
  \alpha(t)
  \sum_{j = 1}^n w_{ij}
  \sum_{r = 1}^n \rho_r
    \, \frac{ h_r(t) } { p_r }
  ,
\end{align*}
%
\note{Derivation.
$$\begin{aligned}
  \dot{ \bar v }
  &
  \stackrel{\mathrm{Eq.\; \eqref{eq:vbar_def}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \sum_{i = 1}^n \rho_i \, \dot v_i
  %
  \stackrel{\mathrm{Eq.\; \eqref{eq:vt_diffeq_mbin}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  %
  \alpha(t)
  \sum_{j = 1}^n \frac{ h_j } { p_j }
                 \sum_{ i = 1 }^{n} \rho_i \, w_{ij}
  \\
  &
  \stackrel{\mathrm{Eq.\; \eqref{eq:w_balance}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t)
  \sum_{j = 1}^n \rho_j
    \, \frac{ h_j } { p_j }
  %
  \stackrel{\mathrm{Eq.\; \eqref{eq:w_sumj}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t)
  \sum_{j = 1}^n w_{ij}
  \sum_{r = 1}^n \rho_r
    \, \frac{ h_r } { p_r }
  .
\end{aligned}$$
}
%
and by Eqs. \eqref{eq:x_def}, \eqref{eq:rhox_sum} and \eqref{eq:h_ave},
%
$$
\begin{aligned}
  \dot x_i(t)
  &= \alpha(t) \sum_{j=1}^n w_{ij}
  \left(
    \frac{ h_j(t) } { p_j }
    -
    \sum_{r = 1}^n
      \rho_r \frac{ h_r(t) } { p_r }
  \right)
  \\
  &\approx
  -\alpha(t) \sum_{j = 1}^n
  w_{ij} \bigl[ x_j(t) - \xi_j (t) \bigr]
  .
\end{aligned}
$$
%where
%we have expanded the right-hand side
%in linear terms of $x_j$
%on the second line.

@@

For simulations with the same $a_0$,
the shape of the mass distribution, $m(Q)$,
also determines how much the optimal schedule
would depend on the simulation length, $T$.
%
Since both $M(Q)$ and $m(Q)$ decrease monotonically with $Q$,
Eq. \eqref{eq:opt_qT} shows that
the optimal $q(T)$ increases with
the simulation length, $T$,
more so if $M(Q)$ has a long tail.
%
Thus, if $M(Q)$ decays slowly because of
many near zero eigenvalues, $\lambda_k$'s (as in the metadynamics case),
both the optimal schedule and
normalized asymptotic error, $\Err_A(T) \cdot T$,
would be sensitive to the simulation length, $T$.
%
In contrast, if all eigenvalues are $1.0$
as in the single-bin or WL updating scheme discussed previously,
the optimal schedule is
given by the universal inverse-time formula,
Eq. \eqref{eq:alpha_invt1},
and the normalized asymptotic error
is roughly a constant
[cf. Eq. \eqref{eq:error_singlebin}].
%
%If $M(Q)$ decays rapidly,
%the distribution $m(Q)$ does not critically
%depend on the cutoff $q(T)$,
%once $q(T)$ is sufficiently large.
%%
%So the optimal schedules of these simulations
%approximately satisfy the scaling relations
%described in Sec. \ref{sec:mass_distr}.
%%
%If, however, $M(Q)$ has a long tail
%because of some near-zero $\lambda_k$'s,
%increasing the simulation length $T$
%generally alters the shape of the optimal schedule.

@@

The analysis in this study has two limitations.
%
First, we have ignored the \emph{systematic}
error\cite{zhou2005, morozov2007, zhou2008}
caused by the updates in adaptive FDS methods.
%
This is valid in the asymptotic regime:
when the updating magnitude is small
and the sampling is a quasi-equilibrium one\cite{
  zhou2005, morozov2007, zhou2008, barducci2008, dama2014},
the random error becomes more important than the systematic one,
because the former is roughly proportional to
the square root of the updating magnitude\cite{
  zhou2005, morozov2007, zhou2008, bussi2006}
whereas the latter is proportional to
the magnitude itself\cite{morozov2007}.
%
%Our method of analysis is most useful for long simulations
%to achieve relatively precise results.
%
Second, by using the white noise approximation,
we have represented the autocorrelation function
of each eigenmode by a single number, $\Gamma_k$.
%
Although this assumption appeared to work well
on our test system for modeling long simulations,
they may break down for relatively short simulations
of glassy systems.
%Our immediate interest is to apply these relations to
%problems containing aqueous mixtures of proteins and nucleic acids.

@@

The above demonstration implies several connections
between the WL algorithm under the optimal inverse-time schedule
with a few related FDS methods.
%
First,
it establishes an efficiency equivalence
between an adaptive WL simulation under the inverse-time schedule
and the ideal equilibrium FDS counterpart
with a nearly perfect bias potential.
%
The bias potentials from the two approaches
would have similar precision at long times.
%
Thus, it is unnecessary to convert
a sufficiently convergent adaptive FDS simulation
to the corresponding equilibirium FDS method
with a fixed bias potential just for better precision
of the final PMF.

Second,
it also relates a WL simulation under the inverse-time schedule
to a time-averaged variant of the regular WL simulation.
%
In the latter, the updating magnitude is fixed,
but the bias potential over the simulation coarse is averaged
as the final result\cite{zhou2005}.
%
The above demonstration shows that the inverse-time schedule
has already effectively implemented an implicit average
so that no explicit averaging is needed in the former approaches.
%
The inverse-time schedule approach is preferred, however,
because with a decreasing updating magnitude,
it minimizes the systematic error caused by
the on-the-fly updating scheme\cite{
  zhou2005, zhou2008},
which is ignored in this analysis.

@@

%A practical problem of using homogeneous updating scheme
%is the following.
%
A homogeneous updating scheme
specified by an arbitrary kernel
is not necessarily stable,
i.e. some eigenvalues can be negative.
%
%
%\note{
  %For example,
  %the triple-bin updating scheme
  %is unstable if Eq. \eqref{eq:nn_stable} is violated.
  %
%}%
Even
the Gaussian updating scheme, which is stable in the continuous limit,
can be unstable upon discretization
during which the updating kernel is truncated.
%
Here we show how to minimally modify
the updating kernel
to stabilize it.
%
\note{The relative code is \texttt{trimwindow()} in \texttt{invt.h}.
}

>>>>

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry
that can provide insight into an array of phenomena not easily studied
with traditional experiments.
%
Given a system,
often the task is to compute
a distribution, $p^*(z)$,
along a collective variable, $z$, by sampling via either Monte Carlo\cite{
  frenkel, newman, landau_binder} (MC)
or molecular dynamics\cite{frenkel, karplus2002} (MD) simulations.
%
The dimensionless free energy, or potential of mean force (PMF),
along the collective variable, $z$,
is $-\ln p^*(z)$.
%
However, accurately estimating the PMF is often difficult
when the collective variable is energetically restricted to local regions
of the complex, multi-modal free energy surface.
%
Thus,
to overcome these energetic barriers and
to capture the global shape of the free energy surface
an effective strategy is to introduce a bias potential that
cancels the PMF.
%
Ideally, then, sampling yields a flat distribution
along $z$\cite{mezei1987, berg1992, *lee1993,
wang2001, *wang2001pre,
huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}.
%
The flat-distribution sampling can be used to accelerate
sampling of slow transitions that otherwise would not readily
be observed with traditional simulations.
%
%A better solution is to artificially flatten
%the target distribution along $z$\cite{mezei1987, berg1992, *lee1993,
%wang2001, *wang2001pre,
%huber1994,
%*laio2002, *laio2008, *barducci2011, *sutto2012}.
%%
%This alleviates the above problem of local trapping
%by forcing the system to travel more
%deliberately along $z$,
%which often represents
%a direction of slow motion.
%%
%To achieve a flat distribution, however,
%the method must be able to construct a
%bias potential that cancels the PMF.

The Wang-Landau (WL) algorithm\cite{wang2001, *wang2001pre}
and metadynamics\cite{huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
flat-distribution-sampling (FDS) techniques
that build the bias potential adaptively or on the fly.
%
Both techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential.
%
%The bias potential is associated with discrete values of $z$.
%
The difference lies
in the updating window function,
which specifies
a neighborhood around the current value of $z$ on the free energy surface
where the updating should occur
as well as the relative updating magnitude.
%
The window function
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one  has to gradually decrease
the updating magnitude of the bias.

@@

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule, $\alpha(t)$,
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014}.
%
An interesting explanation\cite{
  marsili2006, barducci2008}
comes from the interpretation of
the bias potential %under the WL updating scheme, Eq. \eqref{eq:wl_update},
as a runtime average computed from all data points collected so far,
so that the addition of a new data point only
affects the average by a weight of $1/t$,
which is the inverse of the current sample size, $t$
(cf. Appendix \ref{sec:equilerr}).
%
Here we will re-derive this formula
and suggest a slight modification
whereby the origin of $t$ is shifted
according to the initial error.
%
Interestingly, we may reach the same result
from optimally planning the simulation lengths in the stages
and the reduction factor, as shown in Appendix \ref{sec:wltoinvt}.

@@

Note that the distribution, $\pmb \rho$,
introduced above to define the error
is independent of
the target distribution,
$\mathbf p = (p_1, \dots, p_n)$.
%
For example,
in well-tempered metadynamics,
$\pmb\rho$ is a flat distribution,
while $\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
  barducci2008, dama2014} ($\ln p_i = \ln {p^*_i}/(1+\Delta) + \mathrm{const.}$ for some positive $\Delta$).
%
From Eqs. \eqref{eq:x_def} and \eqref{eq:vbar_def},
we have
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%

@@

%
For the random noise part,
the average is zero by definition,
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the autocorrelation functions
depend only on the time difference
%
\begin{equation}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  =
  \sigma_{ij}(t - t'),
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function
that vanishes at large $t$.
%
\note{More explicitly,
if the transition matrix of a period, $\tau$,
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \, \pi_j - \pi_i \, \pi_j.
$
}%
Below, we will assume the noise is white,
such that
$$
\sigma_{ij}(t) = \Gamma_0 \, \delta(t)
$$
for some $\Gamma_0$,
and $\delta(t)$ is the Dirac's delta function.
%
This approximation is valid if the simulation period, $T$,
is much longer than the autocorrelation time of $\sigma_{ij}(t)$.
%
For non-white noise,
the equivalent value of $\Gamma_0$ can be obtained
from the integral of the autocorrelation function,
$\int_{-\infty}^\infty \sigma_{ij}(t) \, dt$.

@@

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
where
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ j = 1 }^n
    \rho_j
    \frac{ \zeta_j(t) }
         {     p_j    }
  ,
\label{eq:xi_def}
\end{equation}
%
is also governed by white noise as is $\zeta_i(t)$,
i.e. we may write
$$
  \langle \xi_i(t) \, \xi_i(t') \rangle = \Gamma_i \, \delta(t - t'),
$$
for some $\Gamma_i$.
%
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ h_j } { p_j }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} \; and \; \eqref{eq:xi_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ \langle h_j \rangle } { p_j }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_ave} \;and\; \eqref{eq:rhox_sum}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

@@

Consider the following stochastic differential equation
of a single variable:
%
\begin{equation}
\dot x(t) = -\alpha(t) \, \left[ x(t) - \xi(t) \right],
\label{eq:dxdt_alpha}
\end{equation}
%
where $\xi(t)$ is some white noise, such that
%
$
  \left\langle \xi(t) \, \xi(t') \right\rangle
  =
  \Gamma \, \delta(t - t')
  .
$
%
We wish to show that the $\alpha(t)$
of minimizing $\langle x^2(t) \rangle$ at long times
is given by
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 },
\label{eq:alpha_invt1}
\end{equation}
%
where $t_0$ is a constant.

@@

We can now simplify Eq. \eqref{eq:vt_diffeq_mbin}.
%
By using Eqs.
\eqref{eq:v_def}, \eqref{eq:vbar_def}, and
\eqref{eq:vt_diffeq_mbin}-\eqref{eq:w_balance}
we get
%
\begin{align*}
  \dot{ \bar v }(t)
  =
  \alpha(t)
  \sum_{r = 1}^n \rho_r
    \, \frac{ h_r(t) } { p_r }
  =
  \alpha(t)
  \sum_{j = 1}^n w_{ij}
  \sum_{r = 1}^n \rho_r
    \, \frac{ h_r(t) } { p_r }
  ,
\end{align*}
%
\note{Derivation.
$$\begin{aligned}
  \dot{ \bar v }
  &
  \stackrel{\mathrm{Eq.\; \eqref{eq:vbar_def}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \sum_{i = 1}^n \rho_i \, \dot v_i
  %
  \stackrel{\mathrm{Eq.\; \eqref{eq:vt_diffeq_mbin}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  %
  \alpha(t)
  \sum_{j = 1}^n \frac{ h_j } { p_j }
                 \sum_{ i = 1 }^{n} \rho_i \, w_{ij}
  \\
  &
  \stackrel{\mathrm{Eq.\; \eqref{eq:w_balance}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t)
  \sum_{j = 1}^n \rho_j
    \, \frac{ h_j } { p_j }
  %
  \stackrel{\mathrm{Eq.\; \eqref{eq:w_sumj}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t)
  \sum_{j = 1}^n w_{ij}
  \sum_{r = 1}^n \rho_r
    \, \frac{ h_r } { p_r }
  .
\end{aligned}$$
}
%
and by Eqs. \eqref{eq:x_def}, \eqref{eq:rhox_sum} and \eqref{eq:h_ave},
%
$$
\begin{aligned}
  \dot x_i(t)
  &= \alpha(t) \sum_{j=1}^n w_{ij}
  \left(
    \frac{ h_j(t) } { p_j }
    -
    \sum_{r = 1}^n
      \rho_r \frac{ h_r(t) } { p_r }
  \right)
  \\
  &\approx
  -\alpha(t) \sum_{j = 1}^n
  w_{ij} \bigl[ x_j(t) - \xi_j (t) \bigr]
  .
\end{aligned}
$$
%where
%we have expanded the right-hand side
%in linear terms of $x_j$
%on the second line.

@@

For simulations with the same $a_0$,
the shape of the mass distribution, $m(Q)$,
also determines how much the optimal schedule
would depend on the simulation length, $T$.
%
Since both $M(Q)$ and $m(Q)$ decrease monotonically with $Q$,
Eq. \eqref{eq:opt_qT} shows that
the optimal $q(T)$ increases with
the simulation length, $T$,
more so if $M(Q)$ has a long tail.
%
Thus, if $M(Q)$ decays slowly because of
many near zero eigenvalues, $\lambda_k$'s,
as in the metadynamics case,
both the optimal schedule and
normalized asymptotic error, $\Err_A(T) \cdot T$,
would be sensitive to the simulation length, $T$.
%
In contrast, if all eigenvalues are $1.0$
as in the single-bin or WL updating scheme discussed previously,
the optimal schedule is
given by the universal inverse-time formula,
Eq. \eqref{eq:alpha_invt1},
and the normalized asymptotic error
is roughly a constant
[cf. Eq. \eqref{eq:error_singlebin}].
%
%If $M(Q)$ decays rapidly,
%the distribution $m(Q)$ does not critically
%depend on the cutoff $q(T)$,
%once $q(T)$ is sufficiently large.
%%
%So the optimal schedules of these simulations
%approximately satisfy the scaling relations
%described in Sec. \ref{sec:mass_distr}.
%%
%If, however, $M(Q)$ has a long tail
%because of some near-zero $\lambda_k$'s,
%increasing the simulation length $T$
%generally alters the shape of the optimal schedule.

@@

The analysis in this study has two limitations.
%
First, we have ignored the \emph{systematic}
error\cite{zhou2005, morozov2007, zhou2008}
caused by the updates in adaptive FDS methods.
%
This is valid in the asymptotic regime:
when the updating magnitude is small
and the sampling is a quasi-equilibrium one\cite{
  zhou2005, morozov2007, zhou2008, barducci2008, dama2014},
the random error,
which is roughly proportional to
the square root of the updating magnitude\cite{
  zhou2005, morozov2007, zhou2008, bussi2006},
can easily outweigh
the systematic error,
which is roughly proportional to
the magnitude itself\cite{morozov2007}.
%
%Our method of analysis is most useful for long simulations
%to achieve relatively precise results.
%
Second, by using the white noise approximation,
we have represented the autocorrelation function
of each eigenmode by a single number, $\Gamma_k$.
%
Although this assumption appeared to work well
on our test system for modeling long simulations,
they may break down for relatively short simulations
of glassy systems.
%Our immediate interest is to apply these relations to
%problems containing aqueous mixtures of proteins and nucleic acids.

@@

The above demonstration implies several connections
between the WL algorithm under the optimal inverse-time schedule
with a few related FDS methods.
%
First,
it establishes an efficiency equivalence
between an adaptive WL simulation under the inverse-time schedule
and the ideal equilibrium FDS counterpart
with a nearly perfect bias potential.
%
The bias potentials from the two approaches
would reach similar precision at long times.
%
Thus, it is unnecessary to convert
a sufficiently convergent adaptive FDS simulation
to the corresponding equilibrium FDS simulation
with a fixed bias potential just for better precision
of the final PMF.

Second,
it also relates a WL simulation under the inverse-time schedule
to a time-averaged variant of the regular WL simulation.
%
In the variant, the updating magnitude is fixed,
but the bias potential over the simulation course is averaged
as the final result\cite{zhou2005}.
%
The above demonstration shows that the inverse-time schedule
has already effectively implemented an implicit average
so that no explicit averaging is needed in the former approach.
%
The inverse-time schedule approach is preferred, however,
because with a decreasing updating magnitude,
it minimizes the systematic error caused by
the on-the-fly updating scheme\cite{
  zhou2005, zhou2008},
which is ignored in this analysis.

@@

%A practical problem of using homogeneous updating scheme
%is the following.
%
A homogeneous updating scheme
specified by an arbitrary kernel
is not necessarily stable,
i.e. some eigenvalues can be negative.
%
%
%\note{
  %For example,
  %the triple-bin updating scheme
  %is unstable if Eq. \eqref{eq:nn_stable} is violated.
  %
%}%
Even
the Gaussian updating scheme, which is stable in the continuous limit,
can be unstable
when the updating kernel is truncated during discretization.
%
Here we show how to minimally modify
the updating kernel
to stabilize it.
%
\note{The relative code is \texttt{trimwindow()} in \texttt{invt.h}.
}
